\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-24 paper/paper.tex
a couple final edits

The cumulative residuals depicted in the two panels of Figure
\ref{fig:residuals} are consistent with our interpretation of the
Hahner twin's Olympic finishing times as not particularly
remarkable. With personal best time as a measure of marathon talent as
in Figure \ref{fig:studentizedresiduals}, the two Hahner residuals are
in the right tail of the residual distribution; however, their
locations are not extreme.  One residual is located at the 87th
percentile and the other at the 92rd.  While these two residuals are
in the tail end of the residual distribution, they are not major
outliers that would lead us to think that the Hahner twin's marathon
finishing times were extremely slow.  Moreover, the residuals for the
Kim twins are similarly not remarkable; these two residuals are in the
left tail of the residual distribution, indicating that the Kims ran
faster than one would have expected.  Finally, if one relies on half
marathon split times as measures of marathon talent, as in Figure
\ref{fig:studentizedresiduals_half}, similar conclusions follows.
Neither Hahner twin had a finishing time that was particularly unusual
given her half marathon split, and this applies to the Kim twins as
well.

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a linear function of the runner's marathon
skill $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. Taking each runner's personal best marathon time as a
proxy for her underlying marathon skill, we estimate $\beta_0$,
$\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect
runner coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

@@

From our simulations we generate intervals which describe the extent
of the variability in marathon finishing times. For example, in 95\%
of the simulations in which Anna Hahner completed the marathon, her
finishing time was between 8,369 seconds and 9,999 seconds. This is
consistent with Anna's observed finishing time in Rio, which was 9,932
seconds. Nonetheless, her finishing time was on the slower end of this interval,.   
In fact, our simulations estimate that, given Anna's personal best marathon time
and her age, she would be expected to finish the Rio marathon in 9,181 seconds, which is slightly more
than 12 minutes faster than her actual time.  This is similar to her sister's result. Lisa Hahner's
corresponding 95\% interval is 8,507 to 10,143 with a mean of 9,319.  Her observed
finishing time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times. And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets). According
to our simulation, then, both Hahner twins did not run appreciably
slowly, conditional on personal best times prior to the Rio Olympics
and age.

>>>>

The cumulative residuals depicted in the two panels of Figure
\ref{fig:residuals} are consistent with our interpretation of the
Hahner twin's Olympic finishing times as not particularly
remarkable. With personal best time as a measure of marathon talent as
in Figure \ref{fig:studentizedresiduals}, the two Hahner residuals are
in the right tail of the residual distribution; however, their
locations are not extreme.  One residual is located at the 87th
percentile and the other at the 92nd.  While these two residuals are
in the tail end of the residual distribution, they are not major
outliers that would lead us to think that the Hahner twin's marathon
finishing times were extremely slow.  Moreover, the residuals for the
Kim twins are similarly not remarkable; these two residuals are in the
left tail of the residual distribution, indicating that the Kims ran
faster than one would have expected.  Finally, if one relies on half
marathon split times as measures of marathon talent, as in Figure
\ref{fig:studentizedresiduals_half}, similar conclusions follows.
Neither Hahner twin had a finishing time that was particularly unusual
given her half marathon split, and this applies to the Kim twins as
well.

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a function of the runner's marathon
skill - which we assume can be determined by a linear combination of her personal best marathon time $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. We estimate $\beta_0$,
$\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect
runner coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

@@

From our simulations we generate intervals which describe the extent
of the variability in marathon finishing times. For example, in 95\%
of the simulations in which Anna Hahner completed the marathon, her
finishing time was between 8,369 seconds and 9,999 seconds. This is
consistent with Anna's observed finishing time in Rio, which was 9,932
seconds. Nonetheless, her finishing time was on the slower end of this interval.   
In fact, our simulations estimate that, given Anna's personal best marathon time
and her age, she would be expected to finish the Rio marathon in 9,181 seconds, which is slightly more
than 12 minutes faster than her actual time.  This is similar to her sister's result. Lisa Hahner's
corresponding 95\% interval is 8,507 to 10,143 with a mean of 9,319.  Her observed
finishing time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times in that their finishes were inside
the bounds of traditional 95\% confidence intervals. And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets). According
to our simulation, then, both Hahner twins did not run appreciably
slowly, conditional on personal best times prior to the Rio Olympics
and age.
